---
title: "Estimation, diagnostics, and uncertainty"
author: "Sven C. van Dijkman"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimation, diagnostics, and uncertainty}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Estimation methods

`nm_est()` exposes thirteen population methods through a common interface.

| Method | Main use |
|---|---|
| FO | Fast first-order approximation |
| FOCE | Conditional estimation without residual interaction |
| FOCEI | Conditional estimation with residual interaction |
| LAPLACE | Conditional mode plus curvature correction |
| ITS | Iterative two-stage estimation |
| GQ | Adaptive generalized Gaussian quadrature marginal likelihood |
| IMP | Importance-sampling marginal likelihood |
| SAEM | Stochastic approximation EM |
| BAYES | Bayesian sampling workflow |
| HMC | Static Hamiltonian Monte Carlo with exact CppAD gradients |
| NUTS | Adaptive no-U-turn Hamiltonian Monte Carlo |
| NPML | Fixed-support nonparametric maximum likelihood |
| NPAG | Adaptive-grid nonparametric maximum likelihood |

Method-specific controls are supplied through `...`. Use a fixed seed for
stochastic methods and record all controls in reports or manifests.

### MU-aware stochastic estimation

When a model declares `MU_n` references, `IMP`, `SAEM`, and `BAYES` use the
mapping by default. The classifier recognizes MU equations that are affine in
native THETAs or in `log(THETA)` coordinates, including subject-level
covariate coefficients. For eligible models:

- IMP re-centres the previous conditional modes as the population MU changes,
  preserving the corresponding individual parameters while finding the next
  proposal;
- SAEM replaces repeated model-level optimization of MU-linked fixed effects
  with a bounded generalized least-squares M-step; and
- BAYES samples the linked fixed effects with a Gaussian MU proposal,
  re-centres ETAs to preserve individual parameters, and applies a
  Metropolis correction so bounds, priors, and any other model dependence
  remain exact.

Nonlinear expressions, inconsistent links, rank-deficient MU designs, direct
use of a linked THETA outside its MU equation, explicit linked-THETA priors,
and currently unsupported random-effect layouts retain the conventional SAEM
path. The exact decision and any runtime fallback are recorded in
`fit$diagnostics$mu_specialization`. Set `mu_specialization = FALSE` as a
reference/debugging control.

```{r mu-estimation-template, eval=FALSE}
saem_fit <- nm_est(
  mu_model, data, method = "SAEM",
  n_iter = 500, mu_specialization = TRUE
)

saem_fit$diagnostics$mu_specialization
```

For GQ, `gq_grid = "auto"` uses a tensor grid through three ETA dimensions and
a Smolyak sparse grid for higher-dimensional models. `gq_order` controls the
tensor nodes per ETA dimension (default 5), while `gq_level` controls the
Smolyak accuracy level (default 3). `gq_adaptive = TRUE` centres and scales the
grid using each subject's conditional mode and exact ETA Hessian, and
`gq_max_points` guards the retained grid size. `gq_adaptive = FALSE` provides
fixed quadrature under OMEGA as a numerical reference. Because Smolyak
combination weights can be negative, the likelihood uses signed log-sum-exp
arithmetic and reports a cancellation ratio. The default CppAD score gradient
treats the finite grid as an integration rule; the slower
`gq_gradient = "finite_grid"` option numerically differentiates the complete
moving-grid approximation. Assess convergence by increasing the selected
grid's order or level.

HMC and NUTS jointly sample population parameters and individual ETAs on an
unconstrained scale. Finite THETA bounds and positive variance parameters are
included through differentiable transforms and their Jacobians. Both methods
adapt step size and a diagonal mass matrix during warmup. NUTS additionally
adapts trajectory length. Retained diagnostics include acceptance, divergent
transitions, NUTS tree-depth saturation, split R-hat, ESS, and per-chain
traces. Increase `target_acceptance` when divergences remain after checking
the model and parameterisation.

NPML optimises mixture weights on a fixed ETA support, initialized from
conditional modes unless `np_supports` is supplied. NPAG repeatedly expands,
reweights, and prunes that support. Both retain support weights, subject-level
posterior support probabilities, and posterior ETA means. Their reported
OMEGA is the weighted support covariance for interoperability, not an assumed
normal random-effect distribution. Support-point inference is non-regular;
use bootstrap rather than the ordinary Hessian covariance step.

```{r gq-template, eval=FALSE}
gq_fit <- nm_est(
  model, data, method = "GQ",
  gq_grid = "auto", gq_order = 5, gq_level = 3,
  covariance = TRUE
)
```

```{r estimation-template, eval=FALSE}
fit <- nm_est(
  model,
  data,
  method = "FOCEI",
  maxit = 200,
  eta_maxit = 100,
  tolerance = 1e-6,
  optimizer_backend = "auto",
  n_cores = 4,
  print_every = 10,
  covariance = TRUE,
  covariance_type = "auto"
)
```

`print_every` writes the objective and scaled population gradient to the
worker log every N outer evaluations. The fit records elapsed model-fit time,
covariance time, iterations, function and gradient evaluations, and available
numerical telemetry.

## Sequential estimation

Several methods can be run as one ordered estimation in the current R process,
local queue, or remote queue. Population estimates initialize the next stage;
compatible conditional methods also receive the previous subject ETA modes.
Only the final stage generates the selected record-level output table, while
every stage retains convergence and timing telemetry.

```{r estimation-sequence, eval=FALSE}
fit <- nm_est_sequence(
  model, data,
  stages = list(
    nm_est_stage("FOCE", maxit = 100, eta_maxit = 50),
    nm_est_stage("SAEM", n_iter = 500, seed = 20260719)
  )
)

fit$method_sequence
fit$stages
```

## Covariance and standard errors

```{r covariance-template, eval=FALSE}
covariance <- nm_cov_step(
  fit,
  type = "auto",
  tolerance = 1e-8,
  samples = 200,
  seed = 20260715
)
```

Available estimators include objective Hessian (`R`), subject-score OPG (`S`),
and robust sandwich covariance. `auto` prefers reliable objective curvature
and falls back when conditioning or positive definiteness is unsatisfactory.
GQ reuses its fitted rule for covariance; IMP and SAEM can use a separate
marginal-information integration budget.

Inspect at least convergence status, gradients, eigenvalues, condition number,
fallback method, standard errors, and relative standard errors before using
the uncertainty estimates.

## Goodness of fit

`nm_gof()` creates record-aligned `PRED`, `IPRED`, residual, and CWRES
quantities. The GUI displays DV versus PRED, DV versus IPRED, CWRES versus
TIME, and CWRES versus PRED. CWRES plots use a -5 to 5 reference range with
horizontal guides at -2 and 2.

```{r gof-template, eval=FALSE}
gof <- nm_gof(fit)
subset(gof, EVID == 0, c(ID, TIME, DV, PRED, IPRED, CWRES))
```

## Predictive diagnostics

```{r predictive-template, eval=FALSE}
vpc <- nm_vpc(
  fit,
  nsim = 500,
  probs = c(0.05, 0.5, 0.95),
  level = 0.90,
  stratify = "SEX",
  seed = 20260715
)

npde <- nm_npde(fit, nsim = 500, seed = 20260715)
npc <- nm_npc(fit, nsim = 500, seed = 20260715)
```

A stratified VPC retains the total-population summary and adds summaries for
each stratum. The GUI generates and saves diagnostic plot data once, then
reloads the saved representation when revisiting a tab.

Binary categorical and time-to-event models have dedicated VPC summaries:

```{r special-vpc-template, eval=FALSE}
categorical_vpc <- nm_vpc_categorical(
  categorical_fit, outcome = "DV", nsim = 500
)

tte_vpc <- nm_vpc_tte(
  tte_fit, event = "EVENT", nsim = 500
)
```

## Bootstrap, SIR, and profile likelihood

```{r uncertainty-template, eval=FALSE}
subject_bootstrap <- nm_bootstrap(
  fit,
  n = 1000,
  strata = "SITE",
  seed = 20260715,
  n_cores = 8
)

cluster_bootstrap <- nm_bootstrap(
  fit, n = 1000, unit = "cluster",
  cluster = "CENTRE", strata = "REGION"
)

parametric_bootstrap <- nm_bootstrap(
  fit, n = 1000, type = "parametric"
)

sir <- nm_sir(
  fit, n_proposal = 5000, n_resample = 2000,
  inflation = 1.5, seed = 20260715
)

profile <- nm_profile(
  fit,
  parameters = c("THETA1", "OMEGA1"),
  points = 11,
  span = 3,
  level = 0.95
)
```

Nonparametric bootstrap samples whole subjects or whole user-declared clusters,
never individual rows, and can preserve the number of units within each
subject/cluster-level stratum. Parametric bootstrap simulates new ETAs and
outcomes from the fitted model before each refit. Every run retains its seed,
resampling design, convergence status, failed-fit messages, parameter samples,
and percentile intervals.

SIR uses a heavy-tailed multivariate proposal derived from the fitted
covariance, evaluates the exact population objective at every proposal, and
reports normalized weights and effective sample size before systematic
resampling. It is useful for irregular or asymmetric frequentist uncertainty;
a very small relative ESS indicates that the proposal or inflation should be
revisited. Profile likelihood fixes one native parameter at each grid point
and re-estimates the others.

## Bayesian predictive checks and comparison

```{r bayesian-diagnostics, eval=FALSE}
log_lik <- nm_log_lik(
  bayes_fit, draws = 500,
  marginal = TRUE, eta_samples = 128
)

ppc <- nm_ppc(
  bayes_fit, draws = 500,
  predictive = "population", stratify = "SEX"
)
waic <- nm_waic(log_lik)
loo <- nm_psis_loo(log_lik)
```

The pointwise unit is an independent subject. By default each posterior
population draw integrates over a new ETA distribution, which is appropriate
for prediction of a held-out subject. Conditional PPCs can instead reuse the
saved subject ETAs for within-subject model checking. PSIS-LOO uses the
optional `loo` package and retains its pointwise Pareto-k diagnostics; inspect
those diagnostics rather than using the total estimate uncritically.

## Reproducible model comparison

```{r comparison-template, eval=FALSE}
comparison <- nm_compare(
  base_fit, covariate_fit,
  labels = c("base", "WT on CL"),
  reference = 1,
  nested = c(TRUE, TRUE),
  diagnostics = list(
    base = list(vpc = base_vpc),
    `WT on CL` = list(vpc = covariate_vpc)
  ),
  bic_n = "subjects",
  parametric_bootstrap = 200,
  seed = 20260715
)

nm_compare_save(comparison, "base-vs-wt.rds")
comparison <- nm_compare_read("base-vs-wt.rds")
```

`nm_compare()` is a durable analysis object rather than only a displayed
table. It retains exact model/data/fit and diagnostic fingerprints, parameter
tables, AIC/BIC and weights, the chosen BIC sample-size convention, comparison
rules, notes, and optional bootstrap-calibrated likelihood-ratio evidence.
Likelihood-ratio inference is produced only when the user explicitly declares
that the candidate nests the reference and the likelihood, data, method, and
degrees of freedom are compatible. Variance-component and mixture additions
carry a boundary warning.

## Stepwise covariate modelling

```{r scm-template, eval=FALSE}
candidates <- data.frame(
  parameter = c("CL", "V"),
  covariate = c("WT", "SEX"),
  form = c("power", "categorical"),
  reference = c(70, NA),
  category = c(NA, 1),
  initial = c(0.75, 0.1)
)

scm <- nm_scm(
  fit,
  candidates,
  direction = "both",
  p_forward = 0.05,
  p_backward = 0.01,
  max_steps = 20
)
```

SCM decisions use nested re-estimation and likelihood-ratio thresholds. Keep
the complete selection log and review biological plausibility rather than
treating automated selection as a substitute for model judgement. The SCM
result contains the same `nm_model_comparison` contract for its base and final
fits, allowing downstream reporting and Help AI to consume the evidence
without reconstructing it.
