---
title: "Estimation, diagnostics, and uncertainty"
author: "Sven C. van Dijkman"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimation, diagnostics, and uncertainty}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Estimation methods

`nm_est()` exposes eight population methods through a common interface.

| Method | Main use |
|---|---|
| FO | Fast first-order approximation |
| FOCE | Conditional estimation without residual interaction |
| FOCEI | Conditional estimation with residual interaction |
| LAPLACE | Conditional mode plus curvature correction |
| ITS | Iterative two-stage estimation |
| IMP | Importance-sampling marginal likelihood |
| SAEM | Stochastic approximation EM |
| BAYES | Bayesian sampling workflow |

Method-specific controls are supplied through `...`. Use a fixed seed for
stochastic methods and record all controls in reports or manifests.

```{r estimation-template, eval=FALSE}
fit <- nm_est(
  model,
  data,
  method = "FOCEI",
  maxit = 200,
  eta_maxit = 100,
  tolerance = 1e-6,
  optimizer_backend = "auto",
  n_cores = 4,
  print_every = 10,
  covariance = TRUE,
  covariance_type = "auto"
)
```

`print_every` writes the objective and scaled population gradient to the
worker log every N outer evaluations. The fit records elapsed model-fit time,
covariance time, iterations, function and gradient evaluations, and available
numerical telemetry.

## Covariance and standard errors

```{r covariance-template, eval=FALSE}
covariance <- nm_cov_step(
  fit,
  type = "auto",
  tolerance = 1e-8,
  samples = 200,
  seed = 20260715
)
```

Available estimators include objective Hessian (`R`), subject-score OPG (`S`),
and robust sandwich covariance. `auto` prefers reliable objective curvature
and falls back when conditioning or positive definiteness is unsatisfactory.
IMP and SAEM can use a separate marginal-information integration budget.

Inspect at least convergence status, gradients, eigenvalues, condition number,
fallback method, standard errors, and relative standard errors before using
the uncertainty estimates.

## Goodness of fit

`nm_gof()` creates record-aligned `PRED`, `IPRED`, residual, and CWRES
quantities. The GUI displays DV versus PRED, DV versus IPRED, CWRES versus
TIME, and CWRES versus PRED. CWRES plots use a -5 to 5 reference range with
horizontal guides at -2 and 2.

```{r gof-template, eval=FALSE}
gof <- nm_gof(fit)
subset(gof, EVID == 0, c(ID, TIME, DV, PRED, IPRED, CWRES))
```

## Predictive diagnostics

```{r predictive-template, eval=FALSE}
vpc <- nm_vpc(
  fit,
  nsim = 500,
  probs = c(0.05, 0.5, 0.95),
  level = 0.90,
  stratify = "SEX",
  seed = 20260715
)

npde <- nm_npde(fit, nsim = 500, seed = 20260715)
npc <- nm_npc(fit, nsim = 500, seed = 20260715)
```

A stratified VPC retains the total-population summary and adds summaries for
each stratum. The GUI generates and saves diagnostic plot data once, then
reloads the saved representation when revisiting a tab.

Binary categorical and time-to-event models have dedicated VPC summaries:

```{r special-vpc-template, eval=FALSE}
categorical_vpc <- nm_vpc_categorical(
  categorical_fit, outcome = "DV", nsim = 500
)

tte_vpc <- nm_vpc_tte(
  tte_fit, event = "EVENT", nsim = 500
)
```

## Bootstrap and profile likelihood

```{r uncertainty-template, eval=FALSE}
bootstrap <- nm_bootstrap(
  fit,
  n = 1000,
  seed = 20260715,
  n_cores = 8
)

profile <- nm_profile(
  fit,
  parameters = c("THETA1", "OMEGA1"),
  points = 11,
  span = 3,
  level = 0.95
)
```

Bootstrap resamples subjects, not rows. Profile likelihood fixes one native
parameter at each grid point and re-estimates the others. These workflows are
particularly useful when Hessian uncertainty is asymmetric or unreliable.

## Stepwise covariate modelling

```{r scm-template, eval=FALSE}
candidates <- data.frame(
  parameter = c("CL", "V"),
  covariate = c("WT", "SEX"),
  form = c("power", "categorical"),
  reference = c(70, NA),
  category = c(NA, 1),
  initial = c(0.75, 0.1)
)

scm <- nm_scm(
  fit,
  candidates,
  direction = "both",
  p_forward = 0.05,
  p_backward = 0.01,
  max_steps = 20
)
```

SCM decisions use nested re-estimation and likelihood-ratio thresholds. Keep
the complete selection log and review biological plausibility rather than
treating automated selection as a substitute for model judgement.
