---
title: "Getting Started with LibeRation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with LibeRation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

LibeRation fits NONMEM-style population PK/PD models using FO, FOCE, FOCEI,
SAEM, Laplace, IMP, and Bayesian MCMC. This vignette walks through a minimal
estimation and goodness-of-fit workflow on synthetic theophylline data.

```{r load}
library(LibeRation)
```

## Synthetic data

Built-in examples mirror common teaching datasets:

```{r catalog}
head(nm_synthetic_catalog()$theo, 3)
```

Generate a small oral two-compartment dataset (ADVAN 4, TRANS 4):

```{r synth}
sim <- nm_synthetic_theo(n_sub = 8L, seed = 1L)
str(sim$model, max.level = 1)
nrow(sim$data$data)
```

## First-order estimation

Start with a fast FO fit using the C++ PK engine:

```{r fo-fit}
fit <- nm_est(
  sim$model,
  sim$data,
  method = "FO",
  grad = "cpp",
  pk_engine = "cpp",
  control = list(
    maxit = 40L,
    compute_inference = FALSE,
    n_cores = 1L
  )
)
fit$method
fit$objective
head(fit$theta, 3)
```

## Goodness-of-fit

Population and individual predictions, residuals, and ETAB summaries:

```{r gof}
pred <- predict(fit, type = "ipred")
head(pred[pred$MDV == 0L, c("ID", "TIME", "DV", "IPRED", "IWRES")])

etab <- nm_etab(fit)
head(etab)
```

## Residual structure configuration

Likelihood/residual models are configured with `nm_lik_config()` and attached
to the model:

```{r lik}
cfg <- nm_lik_config(error = "propadd", omega = "block2")
cfg
```

## Control streams

LibeRation can parse and compose NONMEM control streams for the Shiny editor
and workspace workflow:

```{r ctl}
parts <- nm_ctl_template(
  advan = 4L,
  trans = 4L,
  data_file = "data/theo.csv",
  problem = "THEO example"
)
ctl <- nm_ctl_compose(parts)
grep("\\$PK|\\$THETA|\\$OMEGA", ctl, value = TRUE)
```

Parsed streams round-trip through `nm_ctl_parse()`:

```{r ctl-roundtrip}
parts2 <- nm_ctl_parse(ctl)
nrow(parts2$thetas)
parts2$advan
```

## Project workspace

For Pirana-style project management, initialize a workspace directory:

```{r workspace, eval = FALSE}
ws <- tempfile("LibeRation_ws")
nm_workspace_init(ws, create_demo_project = TRUE)
nm_workspace_list_projects(ws)
# liberation_shiny(workspace = ws)
```

## Background jobs

Long runs can be submitted to a background R process (requires **callr**):

```{r jobs, eval = FALSE}
job <- nm_job_submit(
  sim$model, sim$data,
  method = "FOCE",
  grad = "cpp",
  pk_engine = "cpp",
  control = list(maxit = 100L)
)
job$process$wait()
nm_job_result(job$id)
```

## Next steps

- Estimation methods: `?nm_est`, `?nm_focei_setup`, `?nm_cov_step`
- Simulation: `?nm_simulate`, `?nm_synthetic_catalog`
- GUI: `?liberation_shiny`
- Automatic differentiation engine: **LibeRtAD** (`?LibeRtAD`)
