---
title: "Model specification and numerical solvers"
author: "Sven C. van Dijkman"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model specification and numerical solvers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(LibeRation)
```

## Model blocks

An `nm_model` is serializable and contains model text, parameter tables,
likelihood configuration, solver controls, and compiled expression IR. An
`NMEngine` adds process-local C++ pointers and is rebuilt after serialization.

The principal blocks are:

- `INPUT`: required dataset columns;
- `PRED`: parameter and scaling assignments;
- `ERROR`: observation model;
- `DES`: differential equations for ADVAN6/13;
- `ALG`: algebraic residual equations for experimental index-1 DAEs;
- `THETAS`, `OMEGAS`, and `SIGMAS`: parameter tables; and
- `LIK_CONFIG`: covariance, residual, censoring, IOV, prior, and mixture
  configuration.

Model text can use `THETA(i)`, `ETA(i)`, `SIGMA(i)`, `ERR(i)`, supported
mathematical functions, and tape-safe `ifelse()` expressions.

## Experimental mechanistic and latent-state engines

Longer-range engines are opt-in so that exploratory methods cannot be mistaken
for the established ADVAN paths. Supply
`EXPERIMENTAL = nm_experimental_config(TRUE, label = "purpose")`; this policy
and the exact features used remain in the serialized model and queue payload.

Delay models use `LAG(A(i), delay_name)` in `$DES`, with the delay assignment
defined in `$PK/$PRED`:

```{r dde, eval=FALSE}
model <- nm_model(
  INPUT = c("ID", "TIME", "EVID", "AMT"), ADVAN = 6,
  PRED = "K=THETA(1); TAU=THETA(2); S1=1",
  DES = "DADT(1)=-K*A(1)+0.05*LAG(A(1),TAU)",
  THETAS = data.frame(THETA=1:2, Value=c(0.4, 0.5)),
  DDE_CONFIG = nm_dde_config(step=0.05, minimum_delay=0.5),
  EXPERIMENTAL = nm_experimental_config(TRUE)
)
```

Index-1 DAEs place one `RES(i)` equation per algebraic variable in `$ALG`.
An optional logical sparsity matrix enables independent block solves. Named
stoichiometric systems can instead be generated with `nm_qsp_system()` and
passed to `nm_qsp_model()`.

`nm_factorial_hmm_config()` combines independent latent chains exactly for
tractable joint state spaces. `nm_switching_state_space_config()` jointly
propagates a discrete regime and nonlinear continuous state using the compiled
particle filter and returns filtered and genealogically smoothed regime
probabilities.

`nm_component()` stores immutable hashed dense-network, spline, or
Gaussian-process payloads. `scope="pred"` augments `$PK/$PRED`; `scope="des"`
allows state/time-dependent learned corrections inside `$DES`. Component
execution is offline and part of the C++/CppAD objective.

## Generated output columns

`OUTPUT` selects record-level values to retain with simulations and fitted
runs. `nm_model_outputs()` discovers assignment names from the model text
without executing the model and combines them with standard prediction,
residual, ETA, and compartment outputs.

```{r output-columns, eval=FALSE}
model <- nm_model(
  INPUT = c("ID", "TIME", "EVID", "AMT", "DV", "MDV"),
  OUTPUT = c("PRED", "IPRED", "CWRES", "CL", "V", "A1"),
  ADVAN = 1,
  PRED = "CL=THETA(1)*exp(ETA(1)); V=THETA(2); K=CL/V; S1=V",
  ERROR = "Y=F+ERR(1)",
  THETAS = data.frame(THETA = 1:2, Value = c(2, 20)),
  OMEGAS = data.frame(OMEGA = 1, Value = 0.1),
  SIGMAS = data.frame(SIGMA = 1, Value = 0.1)
)

nm_model_outputs(model)
```

In the GUI, **Validate** compiles the unsaved editor draft and refreshes the
generated-column choices. Selected columns are calculated once with the run,
saved in its result, and merged into Data explorer only when that tab requests
the row payload. The OMEGA table uses `OMEGA(row,col)` labels; ETA names refer
to the random effects governed by that covariance matrix.

Numeric data columns referenced by PRED or DES, such as weight or a laboratory
covariate, are recorded as CppAD dynamic parameters when they do not define the
event topology. Subjects with identical event layout can therefore share one
prediction tape even when those covariates differ. Dose/event fields remain
structural when changing them could alter ordering, ADDL expansion, infusion
handling, or compartment routing. Runtime guards detect changed matrix pivots,
matrix-exponential regimes, adaptive ODE paths, and nonlinear steady-state
convergence paths and automatically retape at a finite valid anchor.

## ADVAN and solver selection

| ADVAN | Structural model | Default propagation |
|---:|---|---|
| 1 | One-compartment IV | analytical |
| 2 | One-compartment with depot | analytical |
| 3 | Two-compartment IV | analytical |
| 4 | Two-compartment with depot | analytical |
| 11 | Three-compartment IV | analytical |
| 12 | Three-compartment with depot | analytical |
| 6 | User ODE system | adaptive Dormand--Prince 5(4) |
| 13 | User ODE system | adaptive implicit trapezoidal |

`SOLVER = "matrix"` selects linear matrix-exponential propagation for a
semantic compartment graph. `SOLVER = "auto"` chooses the specialized ADVAN
kernel when available.

```{r ode-example, eval=FALSE}
ode_model <- nm_model(
  INPUT = c("ID", "TIME", "EVID", "AMT", "CMT", "DV", "MDV"),
  ADVAN = 13,
  DOSECMP = 1,
  OBSCMP = 2,
  PRED = "KFAST=THETA(1); KSLOW=THETA(2); S2=THETA(3)",
  DES = paste(
    "DADT(1) = -KFAST * A(1)",
    "DADT(2) = KFAST * A(1) - KSLOW * A(2)",
    sep = "\n"
  ),
  ERROR = "Y = F + ERR(1)",
  THETAS = data.frame(THETA = 1:3, Value = c(100, 1, 20)),
  SIGMAS = data.frame(SIGMA = 1, Value = 0.1),
  ODE_CONTROL = list(rtol = 1e-7, atol = 1e-9, max_steps = 100000)
)
```

## Parameter bounds and fixed values

THETA rows accept `Lower`, `Value`, `Upper`, and `FIX`. Bounds are applied by
the outer optimizer. Variance parameters must remain non-negative; a zero
OMEGA is allowed only when fixed.

```{r parameter-tables}
theta <- data.frame(
  THETA = 1:2,
  Lower = c(0.01, 1),
  Value = c(2, 20),
  Upper = c(20, 200),
  FIX = c(FALSE, FALSE)
)

omega_full <- data.frame(
  OMEGA = 1:3,
  ROW = c(1, 2, 2),
  COL = c(1, 1, 2),
  Value = c(0.10, -0.02, 0.20)
)
```

A full OMEGA uses the complete lower triangle. Select it with
`nm_lik_config(omega = "full")`; the resulting matrix must be positive
definite. A diagonal table needs only `OMEGA` and `Value`.

## Priors, mixtures, censoring, and IOV

```{r likelihood-configuration}
priors <- rbind(
  nm_prior("THETA1", "normal", mean = 2, sd = 0.5),
  nm_prior("SIGMA1", "lognormal", mean = -2, sd = 0.5),
  nm_prior("OMEGA1", "inverse_gamma", shape = 3, rate = 0.2)
)

likelihood <- nm_lik_config(
  omega = "full",
  sigma_parameterization = "variance",
  blq_method = "m3",
  lloq = 0.05,
  iov = 1,
  occasion_col = "OCC",
  priors = priors,
  mixtures = nm_mixture(c(0.8, 0.2), c("typical", "slow"))
)
```

## First-class continuous, categorical, count, and event outcomes

`nm_outcome()` turns common non-Gaussian and robust observation models into a
declarative part of the model. LibeRation generates an ordinary, editable
`$ERROR` block, compiles it to the C++/CppAD population objective, and retains
the distribution metadata needed for stochastic simulation and diagnostics.

```{r first-class-outcomes, eval=FALSE}
count_model <- nm_model(
  INPUT = c("ID", "TIME", "DV", "MDV"),
  ADVAN = 1,
  PRED = "MU=exp(THETA(1)+ETA(1)); CL=1; V=1; S1=V; F=MU",
  THETAS = data.frame(THETA=1:2, Value=log(c(2, 3)), LOWER=-10, UPPER=10),
  OMEGAS = data.frame(OMEGA=1, Value=0.2),
  OUTCOMES = nm_outcome(
    "negative_binomial", prediction="MU", dispersion="THETA(2)",
    max_count=200
  )
)

simulated <- nm_simulate(count_model, count_design, residual=TRUE, seed=1)
fit <- nm_est(count_model, count_data, method="LAPLACE")
nm_outcome_diagnostics(fit)
nm_vpc_count(fit)
```

Available families are normal, log-normal, fixed-df Student t, Bernoulli,
categorical, ordinal, Poisson, negative binomial, binomial, zero-inflated and
hurdle Poisson, first/recurrent event, competing risks, observed Markov, and an
exact two-state continuous-time Markov model. `nm_irt_outcomes()` creates one
ordinal outcome per item.

Several endpoints can share structural parameters and ETAs in one compiled
likelihood. Give every declaration a unique `dvid` and combine them:

```{r joint-outcomes, eval=FALSE}
joint <- nm_outcomes(
  nm_outcome("normal", name="concentration", dvid=1,
             prediction="CP", scale="SIGMA(1)"),
  nm_outcome("bernoulli", name="response", dvid=2,
             prediction="PRESPONSE")
)
```

The count `max_count` determines how far exact factorial/rising-factorial terms
are unrolled on the AD tape. Data validation rejects a larger observed count.
For event families, each observation row represents an at-risk interval and
`DT` supplies its duration. Use `nm_vpc_tte()`, `nm_vpc_recurrent()`, or
`nm_vpc_competing()` as appropriate.

## Advanced editable structural templates

`nm_model_template()` creates complete ADVAN13 `$PK/$PRED` and `$DES` starts
for nonlinear elimination, transit or dual absorption, parent-metabolite,
effect-compartment, indirect-response, tumour-growth, and TMDD systems. Every
line remains editable and uses the normal ODE/AD engine. Templates that need a
non-zero physiological baseline identify the required initial-state event in
`attr(model, "template_notes")`.

```{r advanced-templates, eval=FALSE}
tmdd <- nm_model_template("tmdd")
transit <- nm_model_template("transit_absorption", n_transit=5)

hazard <- nm_piecewise("TIME", c(24, 72), c("H1", "H2", "H3"))
trend <- nm_spline("TIME", c(0, 12, 48, 96),
                   c("B1", "B2", "B3"), intercept="B0")
```

## User-defined and Markov likelihoods

For categorical, count, survival, Markov, or other non-Gaussian outcomes,
`$ERROR` can provide the observation likelihood directly. Assign either
`LIK`, containing a positive probability or density, or `LOGLIK`, containing
its logarithm. LibeRation compiles the block to C++ expression IR and records
it on the same CppAD tape as prediction, ADVAN/ODE propagation, random effects,
and the population objective.

The following first-order binary Markov model estimates an initial probability
and the 0-to-1 and 1-to-1 transition probabilities. The probabilities are
defined in `$PK/$PRED`, while `$ERROR` selects the contribution associated with
the observed transition.

```{r markov-likelihood, eval=FALSE}
markov <- nm_model(
  INPUT = c("ID", "TIME", "DV", "MDV"),
  ADVAN = 1,
  PRED = "
    PBASE = 1 / (1 + exp(-THETA(1)))
    P01 = 1 / (1 + exp(-(THETA(2) + ETA(1))))
    P11 = 1 / (1 + exp(-(THETA(3) + ETA(1))))
    CL = 1; V = 1; S1 = V; F = PBASE
  ",
  ERROR = "
    PCURRENT = ifelse(
      FIRST == 1,
      ifelse(DV == 1, PBASE, 1 - PBASE),
      ifelse(
        PREV_DV == 0,
        ifelse(DV == 1, P01, 1 - P01),
        ifelse(DV == 1, P11, 1 - P11)
      )
    )
    LOGLIK = log(pmax(PCURRENT, 1e-12))
  ",
  THETAS = data.frame(
    THETA = 1:3, Value = qlogis(c(0.3, 0.2, 0.8)),
    LOWER = -10, UPPER = 10
  ),
  OMEGAS = data.frame(OMEGA = 1, Value = 0.2),
  OUTPUT = c("PBASE", "P01", "P11")
)

fit <- nm_est(markov, transitions, method = "LAPLACE")
```

`PREV_DV`, `PREV_TIME`, `DT`, and `FIRST` are maintained independently for
each subject and `DVID`. A finite outcome on an `MDV = 1` record contributes no
likelihood but still seeds the previous state; this is useful for conditioning
on a known baseline state. `FIRST` is one only when no previous finite outcome
exists in that subject/response sequence.

`LIK`/`LOGLIK` can also use `DV`, `F`/`PRED`/`IPRED`, any assignment created in
`$PK/$PRED`, `THETA()`, `ETA()`, `SIGMA()`, and numeric input columns. Do not
include constants that do not affect estimation unless an absolute normalized
likelihood is required for model comparison. Protect probabilities before
taking logarithms, for example with `pmax(probability, 1e-12)`.

Use `LAPLACE` for the closest NONMEM-style conditional likelihood analysis.
FO, FOCE, and FOCEI are deliberately unavailable because their current
implementations assume a Gaussian residual linearization. ITS, GQ, IMP, SAEM,
Bayesian, HMC/NUTS, and nonparametric methods use the compiled joint likelihood
path. Gaussian WRES/IWRES/CWRES are not defined; use outcome-appropriate
diagnostics. Because a likelihood does not uniquely define an outcome
generator, stochastic simulation must explicitly draw outcomes from the saved
probability outputs.

## Hidden Markov models

For latent-state models, `nm_hmm_config()` connects named `$ERROR`
assignments to a finite-state hidden Markov model. LibeRation evaluates the
initial distribution, transition matrix, and state-conditional emissions for
each record, then combines them with a numerically scaled forward algorithm in
C++. The entire marginal likelihood remains differentiable; hidden state
sequences are not treated as observed outcomes or sampled during deterministic
estimation.

This two-state example models binary observations whose distribution depends
on an unobserved `controlled` or `active` state:

```{r hidden-markov-model, eval=FALSE}
hmm <- nm_model(
  INPUT = c("ID", "TIME", "DV", "MDV"),
  ADVAN = 1,
  PRED = "CL = 1; V = 1; S1 = V; F = 0",
  ERROR = "
    PI_C = 1 / (1 + exp(-THETA(1)))
    PI_A = 1 - PI_C
    P_CC = 1 / (1 + exp(-THETA(2)))
    P_CA = 1 - P_CC
    P_AC = 1 / (1 + exp(-THETA(3)))
    P_AA = 1 - P_AC
    EMIT_C = ifelse(DV == 0, 1 - THETA(4), THETA(4))
    EMIT_A = ifelse(DV == 0, 1 - THETA(5), THETA(5))
  ",
  THETAS = data.frame(
    THETA = 1:5,
    Value = c(qlogis(0.8), qlogis(0.9), qlogis(0.3), 0.1, 0.8),
    LOWER = c(rep(-10, 3), 1e-6, 1e-6),
    UPPER = c(rep(10, 3), 1 - 1e-6, 1 - 1e-6)
  ),
  HMM_CONFIG = nm_hmm_config(
    states = c("controlled", "active"),
    initial = c("PI_C", "PI_A"),
    transition = matrix(
      c("P_CC", "P_CA", "P_AC", "P_AA"), 2, 2, byrow = TRUE
    ),
    emission = c("EMIT_C", "EMIT_A")
  )
)

fit <- nm_est(hmm, observations, method = "LAPLACE")
states <- nm_hmm_decode(fit, method = "all")
gof <- nm_gof(fit) # includes HMM_STATE and HMM_PROB_* columns
```

Rows of `transition` are the state before the transition and columns are the
state after it. Probability-scale initial and transition values are normalized
internally, so they may be probabilities or non-negative weights; setting
`initial_scale = "log"` or `transition_scale = "log"` uses a stable softmax.
Emissions can be positive probabilities/densities or log likelihoods. By
default each `DVID` forms an independent hidden-state sequence within subject;
set `by_dvid = FALSE` to use one sequence. `MDV = 1` records do not emit an
observation or advance the hidden state.

All assignments are evaluated at the current record and may depend on `TIME`,
`DT`, numeric input covariates, `F`, `THETA()`, `ETA()`, and `SIGMA()`. The
transition matrix can therefore be time-inhomogeneous or use a continuous-time
intensity model converted to interval transition probabilities using `DT`.

`nm_hmm_decode()` supports `method = "filtered"`, `"smoothed"`, `"viterbi"`,
or `"all"`. Filtering uses observations up to the current record. The scaled
forward-backward smoother revises each state probability using the complete
sequence, while the log-domain Viterbi decoder returns the single most
probable joint path. `method = "all"` uses explicit `HMM_FILTER_*`,
`HMM_SMOOTH_*`, and `HMM_VITERBI_*` columns; the per-sequence log likelihood,
Viterbi joint log likelihood, and Viterbi log-posterior probability are stored
in `attr(states, "sequence_summary")`. These probabilities condition on the
selected parameter estimates and individual or population ETAs; they do not
integrate parameter uncertainty.

For fitted HMM runs, the workbench adds an **HMM** tab on demand. Its decoder
selector switches among filtered, retrospectively smoothed, Viterbi, and
combined views; subject/sequence and state selectors drive probability and
classified-path plots plus the per-sequence likelihood summary. Decoding rows
are generated only when this tab is first opened for the selected run and are
then cached in the GUI session.

LAPLACE is the recommended deterministic method when random effects are present; GQ, IMP, SAEM,
HMC/NUTS, BAYES, ITS, NPML, and NPAG also use the compiled HMM likelihood.
FO/FOCE/FOCEI remain unavailable because those algorithms assume a Gaussian
residual linearization. Automatic NONMEM control-stream export rejects an HMM
explicitly because NONMEM has no single general control-stream representation
for recursive latent-state likelihoods.

### Continuous-time and observed multi-state models

`nm_cthmm_config()` replaces a discrete transition matrix with an arbitrary
generator. Only off-diagonal rate assignments are supplied; LibeRation creates
each diagonal as the negative row sum and evaluates `exp(Q * DT)` inside the
CppAD objective. The same `nm_hmm_decode()` call then provides filtering,
smoothing, and Viterbi paths for irregularly timed observations. Set
`rate_scale = "log"` when the named `$ERROR` assignments are unconstrained log
rates.

An observed multi-state outcome uses `nm_outcome("continuous_time_markov", ...)`
with matching `initial`, `categories`, and a square off-diagonal `rates`
matrix. Two-state declarations retain their analytical closed form; larger
models are automatically compiled through the general matrix-exponential
sequence engine and support outcome simulation.

### Hidden semi-Markov models

`nm_hsmm_config()` adds an explicit dwell-time distribution to each hidden
state. Rows of its `dwell` matrix are states and columns are durations measured
in observation steps. LibeRation expands this declaration to a sparse
state-duration HMM internally, so the existing exact forward likelihood,
retrospective smoother, Viterbi decoder, and CppAD gradients are reused.
`nm_hmm_decode()` aggregates the expanded representation back to the original
state labels.

```{r hsmm-model, eval=FALSE}
semi_markov <- nm_hsmm_config(
  states = c("controlled", "active"),
  initial = c("PI_C", "PI_A"),
  transition = matrix(c("P_CC", "P_CA", "P_AC", "P_AA"), 2, 2,
                      byrow = TRUE),
  dwell = matrix(c("D_C1", "D_C2", "D_C3",
                   "D_A1", "D_A2", "D_A3"), 2, 3, byrow = TRUE),
  emission = c("EMIT_C", "EMIT_A")
)
```

## Residual covariance

`nm_lik_config()` accepts an estimated AR(1) coefficient through, for example,
`ar1_parameter = "THETA(5)"`. The default `tanh` transform keeps the resulting
correlation strictly between minus one and one. Interleaved DVID observations
retain separate residual histories.

Use `nm_residual_group()` when endpoints observed at the same subject and time
have correlated residuals. Marginal variances continue to come from SIGMA;
the declaration supplies a full DVID correlation matrix. Entries may be fixed
correlations or transformed `THETA(i)`/`SIGMA(i)` references. The exact
conditional likelihood, FO marginal covariance, and stochastic simulator all
use the same block.

`nm_arma_config()` generates an exact linear state-space representation for an
ARMA(p,q) residual process around the structural prediction. AR/MA coefficients
and innovation variance remain ordinary THETA/SIGMA expressions. The generated
`$ERROR` assignments are inserted automatically when `ERROR` is omitted.

## Nested and crossed random effects

`nm_re_config()` maps each ETA block to any dataset grouping column. Blocks
may be nested (study/site/subject) or crossed (subject/reader/device). When no
top-level `cluster` is supplied, LibeRation constructs independent connected
components from all memberships and optimizes the exact joint conditional mode
for each component. Structural PK states still propagate per patient.

```{r random-effect-design, eval=FALSE}
design <- nm_re_config(
  nm_re_block("site", "SITE", 1),
  nm_re_block("patient", "ID", 2:3),
  nm_re_block("reader", "READER", 4)
)

model <- nm_model(
  # ordinary INPUT/PRED/ERROR/parameter declarations ...
  RE_CONFIG = design
)
```

Every ETA must belong to exactly one block. OMEGA correlation is permitted
within a block but not between blocks. A highly connected crossed design can
produce one large conditional mode; supplying a valid independent `cluster`
column limits its dimension.

## Gaussian and nonlinear state-space models

`nm_kalman_config()` connects named `$ERROR` assignments to initial state,
transition, process covariance, observation loading, and observation variance
components. Assignments may use `DT`, covariates, THETA/ETA/SIGMA, or generated
model variables. The compiled Kalman filter contributes its exact marginal
likelihood to estimation; it does not turn latent process states into ETAs.

```{r kalman-model, eval=FALSE}
ou <- nm_model(
  INPUT = c("ID", "TIME", "DV", "MDV"), ADVAN = 1,
  PRED = "CL=1; V=1; S1=V; F=0",
  ERROR = "
    M0=0; P0=THETA(2)
    A11=exp(-THETA(1)*DT)
    Q11=THETA(2)*(1-exp(-2*THETA(1)*DT))
    H1=1; R1=THETA(3)*THETA(3)
  ",
  THETAS = data.frame(THETA=1:3, Value=c(0.3, 0.8, 0.2)),
  KALMAN_CONFIG = nm_kalman_config(
    states = "deviation", initial_mean = "M0",
    initial_covariance = matrix("P0", 1),
    transition = matrix("A11", 1),
    process_covariance = matrix("Q11", 1),
    observation = "H1", observation_variance = "R1",
    by_dvid = FALSE
  )
)

fit <- nm_est(ou, observations, method = "LAPLACE")
states <- nm_kalman_decode(fit)
simulated <- nm_simulate(ou, schedule, residual = TRUE, nsim = 100)
```

The measurement equation defaults to `DV = F + H x + epsilon`; use
`baseline = "zero"` for `DV = H x + epsilon`. `nm_kalman_decode()` reports
one-step predictions, filtered states, innovations, and retrospective
Rauch--Tung--Striebel smoothed states.

For nonlinear dynamics, set `filter = "ekf"`, `"ukf"`, or `"particle"` and
name one next-state assignment per latent state plus a scalar nonlinear
observation assignment. EKF and UKF decoding use a retrospective RTS-style
smoother. The bootstrap particle filter uses saved common random numbers for a
reproducible particle likelihood and returns genealogically smoothed states;
`particles`, `ess_threshold`, and `seed` control its approximation.

Continuous-discrete Itô models use `nm_sde_config()`. Its `drift` vector and
diffusion-loading matrix are `$ERROR` assignments evaluated at fixed substeps.
Euler--Maruyama supports general diffusion; Milstein is available for diagonal
diffusion. EKF/UKF use propagated moments, while the particle filter propagates
stochastic paths directly.

```{r sde-model, eval=FALSE}
sde <- nm_sde_config(
  states = "deviation", initial_mean = "M0",
  initial_covariance = matrix("P0", 1),
  drift = "DRIFT", diffusion = matrix("G", 1),
  observation = "HX", observation_variance = "R",
  filter = "ukf", method = "euler", substeps = 8
)
```

For inter-occasion variability, trailing ETAs are treated as occasion-specific
and the event data must contain the configured occasion column. Mixture models
can refer to `MIXNUM` in `PRED`.

## Infusion and steady state

Event records support `RATE`, `DUR`, `II`, `ADDL`, and `SS`. Analytical ADVAN
models use closed-form periodic steady-state propagation. ODE models use a
periodic shooting/convergence path. Always validate the event sequence and
inspect solver telemetry when combining infusion, steady state, and stiff
systems.

## Restricted C++ model text

`nm_model_cpp()` accepts scalar C++-style assignment expressions with the same
indexed-parameter calls. It deliberately rejects headers, allocation, I/O,
side effects, and arbitrary compiled extensions.

```{r cpp-model, eval=FALSE}
cpp_model <- nm_model_cpp(
  CPP = "CL = THETA(1) * exp(ETA(1)); V = THETA(2); S1 = V;",
  INPUT = c("ID", "TIME", "EVID", "AMT", "CMT", "DV", "MDV"),
  ADVAN = 1,
  ERROR = "Y = F * (1 + ERR(1))",
  THETAS = theta,
  OMEGAS = data.frame(OMEGA = 1, Value = 0.1),
  SIGMAS = data.frame(SIGMA = 1, Value = 0.1)
)
```

Use `nm_support_matrix()` to query implemented feature support in the running
package version.
