---
title: "Model specification and numerical solvers"
author: "Sven C. van Dijkman"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model specification and numerical solvers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(LibeRation)
```

## Model blocks

An `nm_model` is serializable and contains model text, parameter tables,
likelihood configuration, solver controls, and compiled expression IR. An
`NMEngine` adds process-local C++ pointers and is rebuilt after serialization.

The principal blocks are:

- `INPUT`: required dataset columns;
- `PRED`: parameter and scaling assignments;
- `ERROR`: observation model;
- `DES`: differential equations for ADVAN6/13;
- `THETAS`, `OMEGAS`, and `SIGMAS`: parameter tables; and
- `LIK_CONFIG`: covariance, residual, censoring, IOV, prior, and mixture
  configuration.

Model text can use `THETA(i)`, `ETA(i)`, `SIGMA(i)`, `ERR(i)`, supported
mathematical functions, and tape-safe `ifelse()` expressions.

## ADVAN and solver selection

| ADVAN | Structural model | Default propagation |
|---:|---|---|
| 1 | One-compartment IV | analytical |
| 2 | One-compartment with depot | analytical |
| 3 | Two-compartment IV | analytical |
| 4 | Two-compartment with depot | analytical |
| 11 | Three-compartment IV | analytical |
| 12 | Three-compartment with depot | analytical |
| 6 | User ODE system | adaptive Dormand--Prince 5(4) |
| 13 | User ODE system | adaptive implicit trapezoidal |

`SOLVER = "matrix"` selects linear matrix-exponential propagation for a
semantic compartment graph. `SOLVER = "auto"` chooses the specialized ADVAN
kernel when available.

```{r ode-example, eval=FALSE}
ode_model <- nm_model(
  INPUT = c("ID", "TIME", "EVID", "AMT", "CMT", "DV", "MDV"),
  ADVAN = 13,
  DOSECMP = 1,
  OBSCMP = 2,
  PRED = "KFAST=THETA(1); KSLOW=THETA(2); S2=THETA(3)",
  DES = paste(
    "DADT(1) = -KFAST * A(1)",
    "DADT(2) = KFAST * A(1) - KSLOW * A(2)",
    sep = "\n"
  ),
  ERROR = "Y = F + ERR(1)",
  THETAS = data.frame(THETA = 1:3, Value = c(100, 1, 20)),
  SIGMAS = data.frame(SIGMA = 1, Value = 0.1),
  ODE_CONTROL = list(rtol = 1e-7, atol = 1e-9, max_steps = 100000)
)
```

## Parameter bounds and fixed values

THETA rows accept `Lower`, `Value`, `Upper`, and `FIX`. Bounds are applied by
the outer optimizer. Variance parameters must remain non-negative; a zero
OMEGA is allowed only when fixed.

```{r parameter-tables}
theta <- data.frame(
  THETA = 1:2,
  Lower = c(0.01, 1),
  Value = c(2, 20),
  Upper = c(20, 200),
  FIX = c(FALSE, FALSE)
)

omega_full <- data.frame(
  OMEGA = 1:3,
  ROW = c(1, 2, 2),
  COL = c(1, 1, 2),
  Value = c(0.10, -0.02, 0.20)
)
```

A full OMEGA uses the complete lower triangle. Select it with
`nm_lik_config(omega = "full")`; the resulting matrix must be positive
definite. A diagonal table needs only `OMEGA` and `Value`.

## Priors, mixtures, censoring, and IOV

```{r likelihood-configuration}
priors <- rbind(
  nm_prior("THETA1", "normal", mean = 2, sd = 0.5),
  nm_prior("SIGMA1", "lognormal", mean = -2, sd = 0.5),
  nm_prior("OMEGA1", "inverse_gamma", shape = 3, rate = 0.2)
)

likelihood <- nm_lik_config(
  omega = "full",
  sigma_parameterization = "variance",
  blq_method = "m3",
  lloq = 0.05,
  iov = 1,
  occasion_col = "OCC",
  priors = priors,
  mixtures = nm_mixture(c(0.8, 0.2), c("typical", "slow"))
)
```

For inter-occasion variability, trailing ETAs are treated as occasion-specific
and the event data must contain the configured occasion column. Mixture models
can refer to `MIXNUM` in `PRED`.

## Infusion and steady state

Event records support `RATE`, `DUR`, `II`, `ADDL`, and `SS`. Analytical ADVAN
models use closed-form periodic steady-state propagation. ODE models use a
periodic shooting/convergence path. Always validate the event sequence and
inspect solver telemetry when combining infusion, steady state, and stiff
systems.

## Restricted C++ model text

`nm_model_cpp()` accepts scalar C++-style assignment expressions with the same
indexed-parameter calls. It deliberately rejects headers, allocation, I/O,
side effects, and arbitrary compiled extensions.

```{r cpp-model, eval=FALSE}
cpp_model <- nm_model_cpp(
  CPP = "CL = THETA(1) * exp(ETA(1)); V = THETA(2); S1 = V;",
  INPUT = c("ID", "TIME", "EVID", "AMT", "CMT", "DV", "MDV"),
  ADVAN = 1,
  ERROR = "Y = F * (1 + ERR(1))",
  THETAS = theta,
  OMEGAS = data.frame(OMEGA = 1, Value = 0.1),
  SIGMAS = data.frame(SIGMA = 1, Value = 0.1)
)
```

Use `nm_support_matrix()` to query implemented feature support in the running
package version.
