---
title: "Projects, graphical workflow, and NONMEM interoperability"
author: "Sven C. van Dijkman"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Projects, graphical workflow, and NONMEM interoperability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Workspace hierarchy

LibeRation persists work as projects containing named model versions. Each
model version owns numbered estimation or simulation runs, and estimation
runs can own saved GOF, VPC, NPDE, NPC, and covariance results.

```{r workspace}
library(LibeRation)

workspace <- nm_workspace(tempfile("liber-workspace-"))
project <- nm_project_create(
  workspace,
  name = "Oral PK development",
  description = "Example project created from R"
)
nm_project_list(workspace)
```

Model and data objects are serializable. Compiled external pointers are not
stored; they are reconstructed when a version or run is opened.

```{r save-version, eval=FALSE}
version_id <- nm_project_save(
  workspace,
  project$id,
  model = model,
  data = data,
  label = "Mod001"
)

run_id <- nm_project_save_run(
  workspace,
  project$id,
  version = version_id,
  result = fit,
  label = "FOCEI base model"
)

nm_project_save_diagnostics(
  workspace,
  project$id,
  run_id,
  list(gof = nm_gof(fit), vpc = nm_vpc(fit, nsim = 200))
)
```

## Launching the GUI

```{r gui, eval=FALSE}
liber_gui()

# Open an explicit workspace and preload model/data objects.
liber_gui(
  workspace = "D:/projects/LibeR/workspace",
  model = model,
  data = data
)
```

When `workspace` is omitted, Windows uses
`C:/Users/<username>/Documents/LibeR/workspace`; Linux and macOS use
`~/LibeR/workspace`. Client and remote-server settings are stored in user data
locations rather than the installed package directory, so reinstalling the
package does not erase them.

The left panel creates projects and model versions and opens estimation,
simulation, diagnostic, and comparison dialogs. The centre panel shows code,
data, run information, covariance, GOF, and generated predictive diagnostics.
The Jobs tab reconciles persistent local or remote queue states with model
runs.

## Local and remote execution

With LibeRties installed, `liber_gui()` creates a persistent local queue under
the workspace. Remote queues use the same typed job contract. A completed
queue result is attached to its originating model version through stored queue
and job identifiers, preventing duplicate run entries during repeated polls.

See the LibeRties vignettes for service deployment, authentication, resource
limits, cancellation, logs, and reverse-proxy guidance.

## Import a NONMEM control stream

The importer translates the supported semantic subset and preserves
unsupported records in a compatibility report instead of silently discarding
them.

```{r import-control, eval=FALSE}
control <- nm_control_read("run001.ctl", strict = FALSE)
control$model
control$compatibility

fit <- nm_est(control$model, data, method = "FOCEI")
```

Supported records include `$INPUT`, `$DATA`, `$SUBROUTINES`, `$MODEL`,
`$PK`/`$PRED`, `$DES`, `$ERROR`, `$THETA`, diagonal or block `$OMEGA`,
`$SIGMA`, `$ESTIMATION`, `$COVARIANCE`, and `$TABLE`.

## Write a control stream

```{r export-control, eval=FALSE}
text <- nm_control_write(
  model,
  data = "theo.csv IGNORE=@",
  estimation = "METHOD=1 INTERACTION MAXEVAL=9999 PRINT=10",
  covariance = "UNCONDITIONAL"
)
cat(text)

nm_control_write(model, file = "exported.ctl", data = "theo.csv")
```

Round-tripping means that a supported control stream can be imported into the
structured LibeRation representation, edited or run, and exported again while
retaining preserved unsupported text and an explicit compatibility report.
Numerical equivalence should still be checked for the particular model,
dataset, estimation options, and NONMEM version.

## Reports

`nm_report()` creates a deterministic PDF through base R graphics and a JSON
manifest containing numerical summaries and provenance.

```{r report, eval=FALSE}
report <- nm_report(
  fit,
  file = "run001-report.pdf",
  sections = c("summary", "parameters", "gof", "eta", "vpc"),
  vpc = vpc,
  manifest = TRUE
)
```
