---
title: "Projects, graphical workflow, and NONMEM interoperability"
author: "Sven C. van Dijkman"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Projects, graphical workflow, and NONMEM interoperability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Workspace hierarchy

LibeRation persists work as projects containing named model versions. Each
model version owns numbered estimation or simulation runs, and estimation
runs can own saved GOF, VPC, NPDE, NPC, and covariance results.

```{r workspace}
library(LibeRation)

workspace <- nm_workspace(tempfile("liber-workspace-"))
project <- nm_project_create(
  workspace,
  name = "Oral PK development",
  description = "Example project created from R"
)
nm_project_list(workspace)
```

Model and data objects are serializable. Compiled external pointers are not
stored; they are reconstructed when a version or run is opened.

```{r save-version, eval=FALSE}
version_id <- nm_project_save(
  workspace,
  project$id,
  model = model,
  data = data,
  label = "Mod001"
)

run_id <- nm_project_save_run(
  workspace,
  project$id,
  version = version_id,
  result = fit,
  label = "FOCEI base model"
)

nm_project_save_diagnostics(
  workspace,
  project$id,
  run_id,
  list(gof = nm_gof(fit), vpc = nm_vpc(fit, nsim = 200))
)
```

## Launching the GUI

```{r gui, eval=FALSE}
liber_gui()

# Open an explicit workspace and preload model/data objects.
liber_gui(
  workspace = "D:/projects/LibeR/workspace",
  model = model,
  data = data
)
```

When `workspace` is omitted, Windows uses
`C:/Users/<username>/Documents/LibeR/workspace`; Linux and macOS use
`~/LibeR/workspace`. Client and remote-server settings are stored in user data
locations rather than the installed package directory, so reinstalling the
package does not erase them.

The left panel creates projects and model versions and opens estimation,
simulation, diagnostic, and comparison dialogs. The centre panel shows code,
data, run information, covariance, GOF, and generated predictive diagnostics.
The Jobs tab reconciles persistent local or remote queue states with model
runs.

### Visual model builder

The **Visual model** centre tab provides a semantic drag-and-drop editor for
amount and response compartments, directional or bidirectional flows,
clearance, Michaelis--Menten elimination, and arbitrary nonlinear flux
expressions. It previews all generated `$PK`, `$DES`, parameter, and
residual-error code before applying it. Applying never locks the code editor:
the generated model and its graph are retained together, while `$DES` remains
fully editable. Regenerating from the graph is always an explicit action and
warns when the editable code differs from its previous generated baseline.

All visual structural parameters start with log-normal between-subject
variability (`PARAM = THETA(i) * exp(ETA(i))`). IIV can be disabled per
parameter in the visual parameter table.

### Model-definition routes

The Code tab keeps `$PK` and `$PRED` as distinct editable sources and provides
three radio-button routes:

1. **PREDPP / ADVAN (`$PK`)** uses the conventional `$PK -> ADVAN/$DES ->
   $ERROR` workflow.
2. **Direct prediction (`$PRED`)** evaluates each row without compartment
   propagation and requires `$PRED` to assign `F`.
3. **ADVAN + prediction layer (`$PK + $PRED`)** first computes the ADVAN
   prediction, then evaluates a LibeRation post-ADVAN `$PRED` program. It can
   read `F_ADVAN`, `A(i)`, `$PK` assignments, and input covariates and must
   assign the final `F`.

Changing route does not discard either code draft. Validation compiles the
active route, discovers THETA/ETA/SIGMA references across its relevant
programs, and refreshes generated output columns without running the dataset.
The third route is a LibeRation extension; NONMEM export folds the post-ADVAN
program into a marked `$ERROR` block so the numerical order remains explicit.

### Local AI

**Activate AI** is opt-in and remembers the user's consent, selected open
models, backend, and separate Help and Report context settings. WebLLM uses a
dedicated browser WebGPU worker. Ollama uses `ellmer` from the local R session,
discovers models installed at a loopback endpoint, and streams responses
without blocking the Shiny session. The context control is an exact WebLLM
context setting and an input-budget guard for Ollama; Ollama's actual context
is set by its model or Modelfile.

Activation itself does not allocate GPU memory. WebLLM weights are lazy-loaded
and cached only on first use; after initialization, the worker disables its
network APIs before model or run context is sent. Ollama is contacted only on
first generation and is restricted to localhost. It is disabled automatically
for hosted, proxied, session-isolated, or remotely accessed applications;
deployments should additionally call `liber_gui(allow_ollama = FALSE)`.
Neither backend receives LibeRation tools or mutation privileges.

## Local and remote execution

With LibeRties installed, `liber_gui()` creates a persistent local queue under
the workspace. Remote queues use the same typed job contract. A completed
queue result is attached to its originating model version through stored queue
and job identifiers, preventing duplicate run entries during repeated polls.

When the GUI itself is running locally, the Jobs popup can manage a
loopback-only OpenSSH tunnel to a LibeRties service that is reachable from an
SSH destination but is not publicly exposed. The guided setup supports an
optional jump host, an SSH agent or identity file, a temporary connection test,
and automatic reopening during the GUI session. Its readiness assistant can
install the Windows OpenSSH Client, enable `ssh-agent`, discover or generate a
protected key, install its public key on either SSH hop and test the gateway
and destination separately. Passwords, key passphrases and MFA responses stay
inside visible OpenSSH terminals; LibeRation does not capture or store them.
Hosted sessions use direct HTTPS or an externally administered network route
instead.

See the LibeRties vignettes for service deployment, authentication, resource
limits, cancellation, logs, and reverse-proxy guidance.

## Import a NONMEM control stream

The importer translates the supported semantic subset and preserves
unsupported records in a compatibility report instead of silently discarding
them.

```{r import-control, eval=FALSE}
control <- nm_control_read("run001.ctl", strict = FALSE)
control$model
control$compatibility

fit <- nm_est(control$model, data, method = "FOCEI")
```

Supported records include `$INPUT`, `$DATA`, `$SUBROUTINES`, `$MODEL`,
`$PK` and `$PRED`, `$DES`, `$ERROR`, `$THETA`, diagonal or block `$OMEGA`,
`$SIGMA`, `$ESTIMATION`, `$COVARIANCE`, and `$TABLE`. Supported generated
columns in `$TABLE` become the model's `OUTPUT` selection; exporting a model
with selected outputs creates the corresponding `$TABLE` record.
ADVAN5/7 `$MODEL` compartments and Kmn/Km0 rates are translated to an
`nm_matrix_model`. ADVAN9 `$AES`/`$AESINITIAL` blocks are preserved but require
explicit translation to `ALG` and `DAE_CONFIG`; strict import fails rather than
silently dropping an equilibrium constraint.

## Write a control stream

```{r export-control, eval=FALSE}
text <- nm_control_write(
  model,
  data = "theo.csv IGNORE=@",
  estimation = "METHOD=1 INTERACTION MAXEVAL=9999 PRINT=10",
  covariance = "UNCONDITIONAL"
)
cat(text)

nm_control_write(model, file = "exported.ctl", data = "theo.csv")
```

Round-tripping means that a supported control stream can be imported into the
structured LibeRation representation, edited or run, and exported again while
retaining preserved unsupported text and an explicit compatibility report.
Numerical equivalence should still be checked for the particular model,
dataset, estimation options, and NONMEM version.

## Reports

`nm_report()` creates a deterministic PDF through base R graphics and a JSON
manifest containing numerical summaries and provenance.

```{r report, eval=FALSE}
report <- nm_report(
  fit,
  file = "run001-report.pdf",
  sections = c("summary", "parameters", "gof", "eta", "vpc"),
  vpc = vpc,
  manifest = TRUE
)
```

The visual report designer builds a single top-to-bottom workflow from
narrative, run, comparison, diagnostic, and page-break blocks. Narrative
blocks may be user-authored (typed or loaded from TXT/Markdown/DOCX) or drafted
by the browser-local AI from an instruction, template, and source material.
The renderer itself is deterministic and offline: it only consumes the text
already stored in the design and immutable saved-run evidence.

```{r report-design, eval=FALSE}
design <- nm_report_design(list(
  nm_report_block("introduction", text = "Study background"),
  nm_report_block(
    "run", run_ids = run_id,
    elements = c("summary", "parameters", "gof", "vpc", "run_info")
  ),
  nm_report_block("discussion", source = "ai", text = "Reviewed local draft"),
  nm_report_block("conclusion", text = "Conclusions")
))

nm_report_design_save(workspace, project$id, design)
bundle <- nm_report_design_render(
  design, workspace, project$id,
  directory = "reports", formats = c("docx", "pdf")
)
```

DOCX rendering uses Pandoc. PDF rendering uses the R graphics device. Both can
be accompanied by a JSON provenance manifest.
