---
title: "Projects, graphical workflow, and NONMEM interoperability"
author: "Sven C. van Dijkman"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Projects, graphical workflow, and NONMEM interoperability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Workspace hierarchy

LibeRation persists work as projects containing named model versions. Each
model version owns numbered estimation or simulation runs, and estimation
runs can own saved GOF, VPC, NPDE, NPC, and covariance results.

```{r workspace}
library(LibeRation)

workspace <- nm_workspace(tempfile("liber-workspace-"))
project <- nm_project_create(
  workspace,
  name = "Oral PK development",
  description = "Example project created from R"
)
nm_project_list(workspace)
```

Model and data objects are serializable. Compiled external pointers are not
stored; they are reconstructed when a version or run is opened.

```{r save-version, eval=FALSE}
version_id <- nm_project_save(
  workspace,
  project$id,
  model = model,
  data = data,
  label = "Mod001"
)

run_id <- nm_project_save_run(
  workspace,
  project$id,
  version = version_id,
  result = fit,
  label = "FOCEI base model"
)

nm_project_save_diagnostics(
  workspace,
  project$id,
  run_id,
  list(gof = nm_gof(fit), vpc = nm_vpc(fit, nsim = 200))
)
```

## Launching the GUI

```{r gui, eval=FALSE}
liber_gui()

# Open an explicit workspace and preload model/data objects.
liber_gui(
  workspace = "D:/projects/LibeR/workspace",
  model = model,
  data = data
)
```

When `workspace` is omitted, Windows uses
`C:/Users/<username>/Documents/LibeR/workspace`; Linux and macOS use
`~/LibeR/workspace`. Client and remote-server settings are stored in user data
locations rather than the installed package directory, so reinstalling the
package does not erase them.

The left panel creates projects and model versions and opens estimation,
simulation, diagnostic, and comparison dialogs. The centre panel shows code,
data, run information, covariance, GOF, and generated predictive diagnostics.
The Jobs tab reconciles persistent local or remote queue states with model
runs.

### Visual model builder

The **Visual model** centre tab provides a semantic drag-and-drop editor for
amount and response compartments, directional or bidirectional flows,
clearance, Michaelis--Menten elimination, and arbitrary nonlinear flux
expressions. It previews all generated `$PK/$PRED`, `$DES`, parameter, and
residual-error code before applying it. Applying never locks the code editor:
the generated model and its graph are retained together, while `$DES` remains
fully editable. Regenerating from the graph is always an explicit action and
warns when the editable code differs from its previous generated baseline.

All visual structural parameters start with log-normal between-subject
variability (`PARAM = THETA(i) * exp(ETA(i))`). IIV can be disabled per
parameter in the visual parameter table.

### Browser-local AI

**Activate AI** is opt-in and remembers the user's consent, selected open
models, and separate Help and Report context settings. `Auto` selects 8K for
the recommended 3B Help model and a conservative 6K for the 7B Report model;
1K--16K presets and a guarded custom value are also available. The status line
shows the approximate prompt use, output reserve, retained message count, and
whether older context was compacted. If browser GPU allocation fails before
generation, LibeRation retries with a smaller context.

Activation itself does not allocate GPU memory: the dedicated WebGPU
worker and weights are lazy-loaded only when Help or report drafting is first
used. The download is cached under LibeRation's stable local browser origin.
After initialization, the worker disables its network APIs before model or run
context is sent, and receives no tools or DOM access. A compromised browser,
extension, or operating system remains outside that isolation boundary.

## Local and remote execution

With LibeRties installed, `liber_gui()` creates a persistent local queue under
the workspace. Remote queues use the same typed job contract. A completed
queue result is attached to its originating model version through stored queue
and job identifiers, preventing duplicate run entries during repeated polls.

See the LibeRties vignettes for service deployment, authentication, resource
limits, cancellation, logs, and reverse-proxy guidance.

## Import a NONMEM control stream

The importer translates the supported semantic subset and preserves
unsupported records in a compatibility report instead of silently discarding
them.

```{r import-control, eval=FALSE}
control <- nm_control_read("run001.ctl", strict = FALSE)
control$model
control$compatibility

fit <- nm_est(control$model, data, method = "FOCEI")
```

Supported records include `$INPUT`, `$DATA`, `$SUBROUTINES`, `$MODEL`,
`$PK`/`$PRED`, `$DES`, `$ERROR`, `$THETA`, diagonal or block `$OMEGA`,
`$SIGMA`, `$ESTIMATION`, `$COVARIANCE`, and `$TABLE`. Supported generated
columns in `$TABLE` become the model's `OUTPUT` selection; exporting a model
with selected outputs creates the corresponding `$TABLE` record.

## Write a control stream

```{r export-control, eval=FALSE}
text <- nm_control_write(
  model,
  data = "theo.csv IGNORE=@",
  estimation = "METHOD=1 INTERACTION MAXEVAL=9999 PRINT=10",
  covariance = "UNCONDITIONAL"
)
cat(text)

nm_control_write(model, file = "exported.ctl", data = "theo.csv")
```

Round-tripping means that a supported control stream can be imported into the
structured LibeRation representation, edited or run, and exported again while
retaining preserved unsupported text and an explicit compatibility report.
Numerical equivalence should still be checked for the particular model,
dataset, estimation options, and NONMEM version.

## Reports

`nm_report()` creates a deterministic PDF through base R graphics and a JSON
manifest containing numerical summaries and provenance.

```{r report, eval=FALSE}
report <- nm_report(
  fit,
  file = "run001-report.pdf",
  sections = c("summary", "parameters", "gof", "eta", "vpc"),
  vpc = vpc,
  manifest = TRUE
)
```

The visual report designer builds a single top-to-bottom workflow from
narrative, run, comparison, diagnostic, and page-break blocks. Narrative
blocks may be user-authored (typed or loaded from TXT/Markdown/DOCX) or drafted
by the browser-local AI from an instruction, template, and source material.
The renderer itself is deterministic and offline: it only consumes the text
already stored in the design and immutable saved-run evidence.

```{r report-design, eval=FALSE}
design <- nm_report_design(list(
  nm_report_block("introduction", text = "Study background"),
  nm_report_block(
    "run", run_ids = run_id,
    elements = c("summary", "parameters", "gof", "vpc", "run_info")
  ),
  nm_report_block("discussion", source = "ai", text = "Reviewed local draft"),
  nm_report_block("conclusion", text = "Conclusions")
))

nm_report_design_save(workspace, project$id, design)
bundle <- nm_report_design_render(
  design, workspace, project$id,
  directory = "reports", formats = c("docx", "pdf")
)
```

DOCX rendering uses Pandoc. PDF rendering uses the R graphics device. Both can
be accompanied by a JSON provenance manifest.
