---
title: "Estimation, diagnostics, and uncertainty"
author: "Sven C. van Dijkman"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimation, diagnostics, and uncertainty}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Estimation methods

`nm_est()` exposes thirteen population methods through a common interface.

| Method | Main use |
|---|---|
| FO | Fast first-order approximation |
| FOCE | Conditional estimation without residual interaction |
| FOCEI | Conditional estimation with residual interaction |
| LAPLACE | Conditional mode plus curvature correction |
| ITS | Iterative two-stage estimation |
| GQ | Adaptive generalized Gaussian quadrature marginal likelihood |
| IMP | Importance-sampling marginal likelihood |
| SAEM | Stochastic approximation EM |
| BAYES | Bayesian sampling workflow |
| HMC | Static Hamiltonian Monte Carlo with exact CppAD gradients |
| NUTS | Adaptive no-U-turn Hamiltonian Monte Carlo |
| NPML | Fixed-support nonparametric maximum likelihood |
| NPAG | Adaptive-grid nonparametric maximum likelihood |

Method-specific controls are supplied through `...`. Use a fixed seed for
stochastic methods and record all controls in reports or manifests.

For GQ, `gq_grid = "auto"` uses a tensor grid through three ETA dimensions and
a Smolyak sparse grid for higher-dimensional models. `gq_order` controls the
tensor nodes per ETA dimension (default 5), while `gq_level` controls the
Smolyak accuracy level (default 3). `gq_adaptive = TRUE` centres and scales the
grid using each subject's conditional mode and exact ETA Hessian, and
`gq_max_points` guards the retained grid size. `gq_adaptive = FALSE` provides
fixed quadrature under OMEGA as a numerical reference. Because Smolyak
combination weights can be negative, the likelihood uses signed log-sum-exp
arithmetic and reports a cancellation ratio. The default CppAD score gradient
treats the finite grid as an integration rule; the slower
`gq_gradient = "finite_grid"` option numerically differentiates the complete
moving-grid approximation. Assess convergence by increasing the selected
grid's order or level.

HMC and NUTS jointly sample population parameters and individual ETAs on an
unconstrained scale. Finite THETA bounds and positive variance parameters are
included through differentiable transforms and their Jacobians. Both methods
adapt step size and a diagonal mass matrix during warmup. NUTS additionally
adapts trajectory length. Retained diagnostics include acceptance, divergent
transitions, NUTS tree-depth saturation, split R-hat, ESS, and per-chain
traces. Increase `target_acceptance` when divergences remain after checking
the model and parameterisation.

NPML optimises mixture weights on a fixed ETA support, initialized from
conditional modes unless `np_supports` is supplied. NPAG repeatedly expands,
reweights, and prunes that support. Both retain support weights, subject-level
posterior support probabilities, and posterior ETA means. Their reported
OMEGA is the weighted support covariance for interoperability, not an assumed
normal random-effect distribution. Support-point inference is non-regular;
use bootstrap rather than the ordinary Hessian covariance step.

```{r gq-template, eval=FALSE}
gq_fit <- nm_est(
  model, data, method = "GQ",
  gq_grid = "auto", gq_order = 5, gq_level = 3,
  covariance = TRUE
)
```

```{r estimation-template, eval=FALSE}
fit <- nm_est(
  model,
  data,
  method = "FOCEI",
  maxit = 200,
  eta_maxit = 100,
  tolerance = 1e-6,
  optimizer_backend = "auto",
  n_cores = 4,
  print_every = 10,
  covariance = TRUE,
  covariance_type = "auto"
)
```

`print_every` writes the objective and scaled population gradient to the
worker log every N outer evaluations. The fit records elapsed model-fit time,
covariance time, iterations, function and gradient evaluations, and available
numerical telemetry.

## Sequential estimation

Several methods can be run as one ordered estimation in the current R process,
local queue, or remote queue. Population estimates initialize the next stage;
compatible conditional methods also receive the previous subject ETA modes.
Only the final stage generates the selected record-level output table, while
every stage retains convergence and timing telemetry.

```{r estimation-sequence, eval=FALSE}
fit <- nm_est_sequence(
  model, data,
  stages = list(
    nm_est_stage("FOCE", maxit = 100, eta_maxit = 50),
    nm_est_stage("SAEM", n_iter = 500, seed = 20260719)
  )
)

fit$method_sequence
fit$stages
```

## Covariance and standard errors

```{r covariance-template, eval=FALSE}
covariance <- nm_cov_step(
  fit,
  type = "auto",
  tolerance = 1e-8,
  samples = 200,
  seed = 20260715
)
```

Available estimators include objective Hessian (`R`), subject-score OPG (`S`),
and robust sandwich covariance. `auto` prefers reliable objective curvature
and falls back when conditioning or positive definiteness is unsatisfactory.
GQ reuses its fitted rule for covariance; IMP and SAEM can use a separate
marginal-information integration budget.

Inspect at least convergence status, gradients, eigenvalues, condition number,
fallback method, standard errors, and relative standard errors before using
the uncertainty estimates.

## Goodness of fit

`nm_gof()` creates record-aligned `PRED`, `IPRED`, residual, and CWRES
quantities. The GUI displays DV versus PRED, DV versus IPRED, CWRES versus
TIME, and CWRES versus PRED. CWRES plots use a -5 to 5 reference range with
horizontal guides at -2 and 2.

```{r gof-template, eval=FALSE}
gof <- nm_gof(fit)
subset(gof, EVID == 0, c(ID, TIME, DV, PRED, IPRED, CWRES))
```

## Predictive diagnostics

```{r predictive-template, eval=FALSE}
vpc <- nm_vpc(
  fit,
  nsim = 500,
  probs = c(0.05, 0.5, 0.95),
  level = 0.90,
  stratify = "SEX",
  seed = 20260715
)

npde <- nm_npde(fit, nsim = 500, seed = 20260715)
npc <- nm_npc(fit, nsim = 500, seed = 20260715)
```

A stratified VPC retains the total-population summary and adds summaries for
each stratum. The GUI generates and saves diagnostic plot data once, then
reloads the saved representation when revisiting a tab.

Binary categorical and time-to-event models have dedicated VPC summaries:

```{r special-vpc-template, eval=FALSE}
categorical_vpc <- nm_vpc_categorical(
  categorical_fit, outcome = "DV", nsim = 500
)

tte_vpc <- nm_vpc_tte(
  tte_fit, event = "EVENT", nsim = 500
)
```

## Bootstrap and profile likelihood

```{r uncertainty-template, eval=FALSE}
bootstrap <- nm_bootstrap(
  fit,
  n = 1000,
  seed = 20260715,
  n_cores = 8
)

profile <- nm_profile(
  fit,
  parameters = c("THETA1", "OMEGA1"),
  points = 11,
  span = 3,
  level = 0.95
)
```

Bootstrap resamples subjects, not rows. Profile likelihood fixes one native
parameter at each grid point and re-estimates the others. These workflows are
particularly useful when Hessian uncertainty is asymmetric or unreliable.

## Stepwise covariate modelling

```{r scm-template, eval=FALSE}
candidates <- data.frame(
  parameter = c("CL", "V"),
  covariate = c("WT", "SEX"),
  form = c("power", "categorical"),
  reference = c(70, NA),
  category = c(NA, 1),
  initial = c(0.75, 0.1)
)

scm <- nm_scm(
  fit,
  candidates,
  direction = "both",
  p_forward = 0.05,
  p_backward = 0.01,
  max_steps = 20
)
```

SCM decisions use nested re-estimation and likelihood-ratio thresholds. Keep
the complete selection log and review biological plausibility rather than
treating automated selection as a substitute for model judgement.
