---
title: "Getting started with LibeRation"
author: "Sven C. van Dijkman"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with LibeRation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

LibeRation specifies NONMEM-style population PK/PD models in R or restricted
C++ syntax while running model evaluation, event processing, ADVAN/ODE
propagation, likelihood calculations, and automatic differentiation in C++.
This vignette follows a compact oral one-compartment workflow.

## Define a model

Parameter tables use explicit indices and initial values. `Lower`, `Upper`,
and `FIX` columns are optional. Undefined bounds are replaced by the package's
NONMEM-compatible defaults during validation.

```{r model}
library(LibeRation)

model <- nm_model(
  INPUT = c("ID", "TIME", "EVID", "AMT", "CMT", "DV", "MDV"),
  ADVAN = 2,
  TRANS = 2,
  PRED = paste(
    "CL = THETA(1) * exp(ETA(1))",
    "V = THETA(2)",
    "KA = THETA(3)",
    "S2 = V",
    sep = "\n"
  ),
  ERROR = "Y = F * (1 + ERR(1))",
  THETAS = data.frame(
    THETA = 1:3,
    Value = c(2.8, 32, 1.4),
    Lower = c(0.01, 0.1, 0.01),
    Upper = c(20, 200, 10)
  ),
  OMEGAS = data.frame(OMEGA = 1, Value = 0.09),
  SIGMAS = data.frame(SIGMA = 1, Value = 0.15)
)
model
```

`PRED` and `ERROR` retain the established LibeRation syntax, but they are
compiled to expression IR and are not repeatedly evaluated by R.

## Create an event dataset

Rows follow NONMEM event conventions. For ADVAN2, compartment 1 is the depot
and compartment 2 is central.

```{r data}
event_data <- data.frame(
  ID = 1L,
  TIME = c(0, 0.5, 1, 2, 4, 8, 12),
  EVID = c(1L, rep(0L, 6)),
  AMT = c(320, rep(0, 6)),
  CMT = c(1L, rep(2L, 6)),
  DV = NA_real_,
  MDV = c(1L, rep(0L, 6))
)

validated <- nm_dataset(event_data)
validated
```

## Simulate

`nm_simulate()` can return deterministic predictions or sample random effects
and residual error. Replicates can be distributed over persistent PSOCK
workers with `n_cores`.

```{r simulate}
simulated <- nm_simulate(
  model,
  validated,
  nsim = 1,
  random_effects = TRUE,
  residual = TRUE,
  seed = 20260715
)
simulated[c("ID", "TIME", "EVID", "DV", "IPRED")]
```

## Estimate and run covariance

The deterministic approximation estimators are FO, FOCE, FOCEI, Laplace, and
ITS. GQ provides deterministic adaptive Gauss--Hermite integration. IMP,
SAEM, and Bayesian estimation use the same compiled subject evaluators with
method-specific stochastic controls.

```{r estimate, eval=FALSE}
fit <- nm_est(
  model,
  simulated,
  method = "FOCEI",
  maxit = 200,
  eta_maxit = 100,
  covariance = TRUE,
  covariance_type = "auto",
  print_every = 10,
  n_cores = 1
)

summary(fit)
fit$covariance
```

The covariance step returns standard errors, relative standard errors,
correlations, eigenvalues, conditioning diagnostics, and the selected
fallback path. It can also be run later with `nm_cov_step(fit)`.

## Continue to diagnostics or the GUI

```{r next, eval=FALSE}
gof <- nm_gof(fit)
vpc <- nm_vpc(fit, nsim = 200, stratify = "SEX")
npde <- nm_npde(fit, nsim = 200)
npc <- nm_npc(fit, nsim = 200)

liber_gui(model = model, data = simulated)
```

See the other package vignettes for solver selection, uncertainty workflows,
project persistence, queues, and NONMEM control-stream interoperability.
